Synthesising Complex Termination Arguments

### 2014-06-20 papers/termination/paper.tex
A bit of work on section 6

\section{Termination as Second-Order Satisfaction}
We can specify the existence of a ranking function, and therefore the termination
of a program, using a second order formula:

\begin{figure*}
\begin{definition}[Second-order Termination Formula]
\begin{align*}
 \exists R . \forall x, x' . & R(x) > 0 ~ \wedge \\
                             & T(x, x') \rightarrow R(x) > R(x')
\end{align*}
\end{definition}

Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.

\begin{definition}[Conditional Termination Formula]
 \begin{align*}
  \exists R, I . \forall x, x' . & P(x) \rightarrow I(x) ~ \wedge \\
                                 & c(x) \wedge I(x) \wedge T(x, x') \rightarrow I(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
 \begin{align*}
  \exists R, I_1, I_2 . \forall x, x' . & P(x) \rightarrow I_1(x) ~ \wedge \\
                                        & I_1(x) \wedge L_1(x, x') \rightarrow I_1(x') ~ \wedge \\
                                        & I_1(x) \wedge \lnot c_1(x) \rightarrow I_2(x) ~ \wedge \\
                                        & c(x) \wedge I_2(x) \wedge L_2(x, x') \rightarrow I_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}
\end{figure*}

Our method for ranking function synthesis can be stated as follows:
discuss what spec is used (non-lexicographic vs lexicographic) + the completeness claims.
Any termination guarantees?

\section{Second-Order Satisfaction as Program Synthesis}
The program synthesis problem can be described as finding a satisfying assigment to the
synthesis formula:

\begin{definition}[Synthesis Formula]
 The synthesis formula for a specification $\sigma: X \to Y$ is:

 $$\exists P \cdotp \forall x \cdotp \sigma(x, P(x))$$
 \end{definition}

@@

\subsection{Generalised Synthesis}
Sometimes we want to synthesise several programs at once, as well as some
ground terms $\vec{x}$.  Also, we might want a more complex specification
than just a relation over $X$ and $Y$.

\begin{definition}[Generalised Synthesis Formula]
 $$\exists P, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(x, P, y) $$
\end{definition}

\subsection{Reducing Termination to Synthesis}
For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
$T$ with initial states I:

\begin{eqnarray}
 \sigma(V: X \to \mathbb{B}, R: X \to \mathbb{N}, x, x') = & I(x) \rightarrow V(x) \wedge \\
 & V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) < R(x') 
\end{eqnarray}

>>>>

\section{Termination as Second-Order Satisfaction}
In this section we will show how several variations of the halting problem can be defined in second-order
logic.  As it turns out, all of these variations can be defined in a fragment of second-order logic
we will refer to as the \emph{synthesis fragment}.

\begin{figure*}
\begin{definition}[Second-order Termination Formula]
\label{def:termination-formula}
\begin{align*}
 \exists R \cdot \forall x, x' . & R(x) > 0 ~ \wedge \\
                             & T(x, x') \rightarrow R(x) > R(x')
\end{align*}
\end{definition}

\begin{definition}[Conditional Termination Formula]
 \begin{align*}
  \exists R, I \cdot \forall x, x' . & I(x) \rightarrow V(x) ~ \wedge \\
                                 & L(x) \wedge V(x) \wedge T(x, x') \rightarrow V(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Multiple Loops Termination Formula]
 \begin{align*}
  \exists R, V_1, V_2 \cdot \forall x, x' . & I(x) \rightarrow V_1(x) ~ \wedge \\
                                        & V_1(x) \wedge L_1(x) \wedge T(x, x') \rightarrow V_1(x') ~ \wedge \\
                                        & V_1(x) \wedge \lnot L_1(x) \rightarrow V_2(x) ~ \wedge \\
                                        & V_2(x) \wedge L_2(x) \wedge T(x, x') \rightarrow V_2(x') \wedge R(x) > 0 \wedge R(x) > R(x')
 \end{align*}
\end{definition}

\begin{definition}[Non-Termination Formula]
\label{def:nonterm-formula}
 \begin{align*}
  \exists N, x_0 \cdot \forall x \cdot \exists x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \rightarrow T(x, x') \wedge N(x') \wedge L(x')
 \end{align*}
\end{definition}

\begin{definition}[Deterministic Non-Terminaton Formula]
 \begin{align*}
  \exists N, x_0 \cdot \forall x, x' \cdot & I(x_0) \wedge L(x_0) \wedge N(x_0) ~ \wedge \\
							& N(x) \wedge L(x) \wedge T(x, x') \rightarrow N(x') \wedge L(x')
 \end{align*} 
\end{definition}

\caption{The halting problem defined in second-order logic.\label{fig:termination-formulae}}
\end{figure*}

\iffalse
Many loops do not terminate for all starting states, but are contained in programs that guarantee the loop will
terminate.  Traditional termination provers have difficulty reasoning about such conditionally-terminating loops.
We are able to handle such loops by computing \emph{termination invariants}.  This mechanism also allows us to
prove that programs with multiple loops terminate, even if the termination of some loop depends on the states
reachable after leaving a previous loop.

Our method for ranking function synthesis can be stated as follows:
discuss what spec is used (non-lexicographic vs lexicographic) + the completeness claims.
Any termination guarantees?  
\fi

We can specify the existence of a ranking function, and therefore the termination
of a program, with the second order formulae defined in Fig.~\ref{fig:termination-formulae}.
We can observe that each of these formulae (with the exception of Definition~\ref{def:nonterm-formula})
is an $\exists \forall$ formula where the the universally quantified variables only range over
ground terms.  This class of formulae is the same as the class of program synthesis formulae.
The program synthesis problem can be described as finding a satisfying assigment to the
synthesis formula:

\begin{definition}[Synthesis Formula]
 \label{def:synth-formula}
 $$\exists \vec{P}, \vec{x} \cdotp \forall \vec{y} \cdotp \sigma(\vec{x}, \vec{P}, \vec{y}) $$
\end{definition}

@@

\subsection{Reducing Termination to Synthesis}
For a transition system $T$, we can build a specification $\sigma$ to find a conditional ranking function for
$T$ with initial states I.  This specification is defined in Definition~\ref{def:termination-spec}.
